---
Title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

### General Information
 * message=FALSE and warning=FALSE are used to suppress warnings and mesages.  
 * Total Number of steps are plotted 3 ways
    * Histogram including NAs.
    * Histogram excluding NAs.
    * Time Series plot replacing NAs with mean for corresponding intervals.
 * NAs are excluded from calculation of Average steps per interval as they cause the averages to be NA.
        
## Loading and preprocessing the data
Download and unzip file
```{r, echo=TRUE, message=FALSE, warning=FALSE}
    library(dplyr)
    URL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(URL,destfile="activity.zip")
    unzip("activity.zip")
```
Read file using read.csv() into a dataframe and check dataframe summary.
```{r, echo=TRUE}
    activity <- read.csv("activity.csv")
    summary(activity)
```
## What is mean total number of steps taken per day?
Calculate total number of steps per day using dplyr package. Total number of steps are calculated in two ways - including NAs and excluding NAs.
Include NAs in total calculation
```{r, echo=TRUE,fig.height=6,fig.width=6,message=FALSE, warning=FALSE}
    total_steps_per_day <- activity %>%
                                  group_by(date) %>%
                                      summarize(total=sum(steps))
    hist(total_steps_per_day$total,xlab="Total Number of Steps Each Day", main="Histogram - Total Number of Steps Per Day - Inlcude NA",col="cyan")
    mean_include_NA <- mean(total_steps_per_day$total)  
    median_include_NA <- median(total_steps_per_day$total)  
```
Mean of Total Number of Steps Per Day : `r mean_include_NA`  
Median of Total Number of Steps Per Day : `r median_include_NA`    

Exclude NAs to avoid summarizing to NA, and to avoid setting mean and meadian to NA.
```{r, echo=TRUE,fig.height=6,fig.width=6,message=FALSE, warning=FALSE}
    total_steps_per_day <- activity[!is.na(activity$steps),] %>%
                                  group_by(date) %>%
                                      summarize(total=sum(steps))
    hist(total_steps_per_day$total,xlab="Total Number of Steps Each Day", main="Histogram - Total Number of Steps Per Day - Exclude NA",col="cyan")
    mean_exclude_NA <- mean(total_steps_per_day$total)  
    median_exclude_NA <- median(total_steps_per_day$total)  
```
Mean of Total Number of Steps Per Day : `r mean_exclude_NA`  
Median of Total Number of Steps Per Day : `r median_exclude_NA`    

## What is the average daily activity pattern?
Caluclate mean/average for each interval using summarize function. Also, calculate max average and corresponding interval and mark on the plot.
```{r, echo=TRUE,message=FALSE, warning=FALSE}
    average_steps <- activity[!is.na(activity$steps),] %>%
                          group_by(interval) %>%
                                summarize(average=mean(steps))
    max_average <- max(average_steps$average)
    interval_max_average <- average_steps[average_steps$average==max_average,1]
    plot(average_steps$interval,average_steps$average,type = "l",ylab="Average Steps", xlab= "Interval",main = "Average Steps Per Interval", col="cyan")
    points(interval_max_average,max_average,col="Red")
   # Plot maximum average using a red point.
```
Interval `r interval_max_average` has the maximum avarage of `r max_average`. Plotted using a red point.  

## Imputing missing values. 
Number of Rows with missing values = `r sum(!complete.cases(activity))`  
Replace missing values with mean/average for the corresponding interval. 
```{r, echo=TRUE, message=FALSE, warning=FALSE}
    for (int in unique(activity[,3]))
    {
        activity[is.na(activity$steps) & activity$interval==int,1] <-  average_steps[average_steps$interval==int,2]   
    }
```
Number of Rows with missing values after replacement = `r sum(!complete.cases(activity))`  

Calculate total number of steps per day using dplyr package. NAs are replaced with mean of corresponding interval. 
```{r, echo=TRUE,fig.height=6,fig.width=6, message=FALSE, warning=FALSE}
    total_steps_per_day_wo_NA <- activity %>%
                                      group_by(date) %>%
                                          summarize(total=sum(steps))
    hist(total_steps_per_day_wo_NA$total,xlab="Total Number of Steps Each Day", main="Histogram - Total Number of Steps Per Day - Replace NA",col="cyan")
    mean_wo_NA <- mean(total_steps_per_day_wo_NA$total)
    median_wo_NA <- median(total_steps_per_day_wo_NA$total)
```
Mean of Total Number of Steps Per Day with NA replaced : `r mean_wo_NA`  
Median of Total Number of Steps Per Day with NA replaced : `r median_wo_NA`    

Comparison of mean and median without any action on NAs, excluding NAs and replacing NAs with mean for corresponding interval.

Handling Scenario     | Mean  | Median
----------------------|-------|------
No Action             | `r mean_include_NA` | `r median_include_NA` 
Exclude NA            | `r mean_exclude_NA` | `r median_exclude_NA` 
Replace NA with Mean  | `r mean_wo_NA` | `r median_wo_NA`

## Are there differences in activity patterns between weekdays and weekends?
Use lattice plot to plot the activity pattern for weekends and weekdays. Added a new column to activity data set to indicate whether a date is weekend or weekday.
```{r, echo=TRUE,, message=FALSE, warning=FALSE}
    library(lubridate)
    library(lattice)
    activity$weekday_end <-ifelse(weekdays(ymd(activity$date)) %in% c("Saturday","Sunday"),"weekend","weekday")
    average_steps <- activity %>%
                        group_by(weekday_end,interval) %>%
                            summarize(average=mean(steps))        
   # average_steps <- transform(average_steps, weekday_end=factor(weeday_end))
    xyplot(average ~ interval | weekday_end,
           data=average_steps,
           type="l",
           layout=c(1,2),
           xlab="Interval",
           ylab="Number of Steps",
           col="cyan")
```
